---
title: "Analysis of Miles per Gallon for Manual and Automatic Transmission"
author: "Ivan Lysiuchenko"
date: "June 23, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Executive summary


## Initial exploratory data analysis

```{r echo=FALSE, message=FALSE}
library(dplyr)
library(GGally)
cardata <- mtcars
cardata <- mutate(cardata, 
                  cyl = factor(as.character(cyl)),
                  vs = factor(as.character(vs)),
                  am = factor(as.character(am)),
                  carb = factor(as.character(carb))
                  )
meanAuto <- mean(cardata[cardata$am == 0, "mpg"])
meanManual <- mean(cardata[cardata$am == 1, "mpg"])
```

Let's start taking a look at interactions between the predictors and outcome.
One of the panels present in the pairs plot (Figure 1 of the Appendix) and a direct
calculation show that really there is a better mean miles per gallon (mpg)
value for manual transmission (`r meanManual`) than for auto (`r meanAuto`).

Other aspects we notice are some patterns looking like linear relationships,
between the outcome (mpg) and some variables, e.g. disp, hp, wt.

## Search for a reasonable model

Some variables take discrete values, so we treat them as factor variables:
cyl, vs, am, gear, carb. The cardata structure mentioned in the code chunks
that follow is the same as mtcars, but the mentioned columns have been converted 
into factors.

If we just fit the mpg as a function of am, we don't get anything interesting.
The estimates would be the mean mpg values for the two groups of cars formed
by differentiating their transmission type.Maybe, we can get some insight looking 
at a complete model, where all the variables are used as predictors:

```{r echo=TRUE, eval=FALSE}
modelComplete <- lm(data = cardata, formula = mpg ~ .)
```

Analyzing the pvalues, we can say that there are no coefficients significant
with Type I error probability of 0.05. Only two of them, hp (horsepower)
and wt (weight) are somehow close to be significant. Let's take one of them 
to include in the model we fitted for am as a predictor:

```{r}
modelAmWt <- lm(data = cardata, formula = mpg ~ am + wt)
summary(modelAmWt)$coefficients
```

It turns out that the inclusion of weight makes the coefficient for am insignificant
with a very high p-value. Let's look at a plot showing how mpg depends on wt. We'll
use color to mark data for automatic (0) and manual (1) transmission.

```{r}
g <- ggplot(data.frame(wt = cardata$wt, mpg = cardata$mpg, am = cardata$am), 
            aes(x = wt, y = mpg, color = am)) + geom_point() + 
    labs(x = "weight, 1000 lbs", y = "miles per gallon",
         title = "Miles per gallon as a function of weight")
```

It follows that according to our data, automatic transmission is generally 
installed on heavier cars. Therefore, wt appears as a confounding variable rendering
insignificant the difference based on the 'auto'/'manual' grouping.

Figure 2 in the Appendix shows that generally for the same horsepower, lighter cars
have manual transmission and the heavier ones automatic. So, for a given horsepower,
wt and am explain roughly the same thing. We can fit a model using am and hp.
All its coefficients are significant.

```{r echo=FALSE}
modelAmHp <- lm(data = cardata, formula = mpg ~ am + hp)
summary(modelAmHp)$coefficients
```

Let's interpret this model. With the same type of transmission, when the horsepower
is increased by 1 unit, the miles per gallon change by `r modelAmHp$coefficients[3]`
(decrease).
Fixed the horsepower, when we change from automatic transmission to manual,
the miles per gallon change by `r modelAmHp$coefficients[2]` (increase).

But once more, if wt is included, am loses its statistical significance. 
Moreover, it seems more logical to assume that the automatic transmission is
installed on heavier cars because they are heavier, the opposite claim 
("a car is heavier because its transmission is automatic") has no sense. In our opinion, it's better to take weight and horsepower as predictors and try 
to build a better model.

## Building a model




## Appendix

```{r message=FALSE, echo=FALSE}
#library(GGally)
#ggpairs(mtcars, upper = list(continuous = "points", combo="dot_no_facet")) + 
#    labs(title = "Pairs plot")
```


```{r}


# Figure 2
#g <- ggplot(data.frame(hp = cardata$hp, mpg = cardata$mpg, am = cardata$am), 
#            aes(x = hp, y = mpg, color = am)) + geom_point() + 
#    labs(x = "horsepower", y = "weight, 1000 lbs", 
#         title = "Figure 2. Weight and horsepower")

#g <- ggplot(data.frame(wt = cardata$wt, hp = cardata$hp, am = cardata$am), aes(x = hp, 
#y = wt, color = am)) + geom_point()



#g <- ggplot(data = data.frame(mpg = cardata$mpg, residWtHp = resid(modelWtHp), residWtHpInteraction = resid(modelWtHpInteraction), residAmHp = resid(modelAmHp))) + geom_line(aes(x = mpg, y = residWtHp), color = "blue") + geom_line(aes(x = mpg, y = residWtHpInteraction), color = "red") + geom_line(aes(x = mpg, y = residAmHp), color = "green")

```


```{r}
#modelAm <- lm(data = cardata, formula = mpg ~ am)






#g <- ggplot(data.frame(x = cardata$mpg, y = resid(modelWtHpInteraction), am = 
#cardata$am), aes(x = x, y = y, color = am)) + geom_point()
```